---
output: github_document
bibliography: "./vignettes/references.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# powRICLPM

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/powRICLPM)](https://CRAN.R-project.org/package=powRICLPM)
[![R-CMD-check](https://github.com/JeroenDMulder/powRICLPM/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JeroenDMulder/powRICLPM/actions/workflows/R-CMD-check.yaml)
 <a href="https://www.r-pkg.org/pkg/powRICLPM"><img src="https://cranlogs.r-pkg.org/badges/grand-total/powRICLPM" alt="CRAN RStudio mirror downloads"/></a>
<!-- badges: end -->


`powRICLPM` is an R package that aids researchers with performing power analyses for the random intercept cross-lagged panel model (RI-CLPM) by @hamaker_critique_2015, the Stable Trait Autoregressive Trait State Model (STARTS) by @kenny_trait-state-error_1995 and @kenny_trait-state_2001, and the dynamic panel model (DPM). It implements the strategy as proposed by @mulder_power_2023. Its main functionalities include:

- [**Basic power analysis**](https://jeroendmulder.github.io/powRICLPM/articles/start.html): Use Monte Carlo simulations to compute p-value based power to reject the null hypothesis (as well as other performance measures such as bias and mean square error) for model parameters in the RI-CLPM, STARTS, and DPM, for a specific experimental condition. Conditions can vary by sample size, number of repeated measures, RI-CLPM/STARTS between-unit variance proportion or DPM accumulating-factor proportion, and reliability when measurement error is included. `powRICLPM` can perform power analyses across multiple experimental conditions simultaneously, and report the results back in a user-friendly manner.
- [**Extensions**](https://jeroendmulder.github.io/powRICLPM/articles/extensions.html): The basic power analysis setup can be extended to include bounded estimation, measurement error, reliability conditions, time-varying loadings, various (stationarity) constraints over time on parameters of the estimation model, DPM accumulating-factor constraints, and the generation of nonnormal data, among other things.
- [**Mplus**](https://jeroendmulder.github.io/powRICLPM/articles/mplus.html): When Mplus is installed, `powRICLPM` can create Mplus syntax, and run the power analyses in Mplus. 

## Documentation

There are four sources of documentation for `powRICLPM`:

- The rationale for the power analysis strategy underlying this package can be found in @mulder_power_2023.
- Every user-facing function in the package is documented, and the documentation can be accessed by running `?function_name` in the R console (e.g., `?powRICLPM`). Here, you can find explanations on how to use the functions, as well as technical details.
- More elaborate descriptions of this package's functionality and analysis options are described in vignettes. These are accessible via the 'Vignettes' tab in the menu, or via R using `vignette(package = "powRICLPM")`. 
- The [FAQ](https://jeroendmulder.github.io/powRICLPM/articles/FAQ.html) contains answers to frequently asked question that reach me via email.

## Installation

To install the development version of `powRICLPM`, including the latest bug fixes and new features, run:

```{r gh-installation, eval = FALSE}
install.packages("devtools")
devtools::install_github("jeroendmulder/powRICLPM")
```

To install the latest release of `powRICLPM` from CRAN, run:

```{r gh-install-CRAN, eval = FALSE}
install.packages("powRICLPM")
```


## Citing `powRICLPM`

You can cite the R-package with the following citation:

> Mulder, J.D., (2023). Power analysis for the random intercept cross-lagged panel model using the powRICLPM R-package. *Structural Equation Modeling: A Multidisciplinary Journal, 30*(4), 645-658. https://doi.org/10.1080/10705511.2022.2122467 

## Contact

If you have ideas, comments, or issues you would like to raise, please get in touch.  

* Issues and ideas can be raised on GitHub via [https://github.com/jeroendmulder/powRICLPM](https://github.com/jeroendmulder/powRICLPM)
* Pull request can be created on GitHub via [https://github.com/jeroendmulder/powRICLPM/pulls](https://github.com/jeroendmulder/powRICLPM/pulls)

## References
